---
title: "Factors that correlate to the spread of COVID-19"
author: "John Lee"
date: '2021 5 6 '
output: html_document
---
# 1. Data setup
```{r}
library(mgcv) # for GAM function
library(car) # for vif function
# The code will only work after wd is set to the directory with the csv files
wd = "/Users/johnlee/Downloads/Factors that correlate to the spread of COVID-19/processed data/"
# dat1 includes countries with and without testing data, and consists of 164 data points
dat1 = read.csv(paste(wd, "processed data.csv", sep = ""))
# dat2 includes only countries with testing data, and consists of 104 data points
dat2 = read.csv(paste(wd, "processed data with testing.csv", sep = ""))
```

# 2. Exploratory Data Analysis

## 2.1. all variables VS Cases_per_mill, dat1 on top, dat2 on bottom
```{r}
par(mfrow=c(2,3))
plot(dat1$Population, dat1$Cases_per_mill)
plot(dat1$Population_Density, dat1$Cases_per_mill)
plot(dat1$GDP_Per_Cap, dat1$Cases_per_mill)
plot(dat2$Population, dat2$Cases_per_mill)
plot(dat2$Population_Density, dat2$Cases_per_mill)
plot(dat2$GDP_Per_Cap, dat2$Cases_per_mill)

plot(dat1$Resp_Death_Rate, dat1$Cases_per_mill)
plot(dat1$Median_Age, dat1$Cases_per_mill)
plot(dat1$Contact_Tracing, dat1$Cases_per_mill)
plot(dat2$Resp_Death_Rate, dat2$Cases_per_mill)
plot(dat2$Median_Age, dat2$Cases_per_mill)
plot(dat2$Contact_Tracing, dat2$Cases_per_mill)

plot(dat1$INT_Travel_Controls, dat1$Cases_per_mill)
plot(dat1$Physicians_Per_1000_ppl, dat1$Cases_per_mill)
plot(dat1$Percent_Urbanization, dat1$Cases_per_mill)
plot(dat2$INT_Travel_Controls, dat2$Cases_per_mill)
plot(dat2$Physicians_Per_1000_ppl, dat2$Cases_per_mill)
plot(dat2$Percent_Urbanization, dat2$Cases_per_mill)

plot(dat1$Life_Expectancy, dat1$Cases_per_mill)
plot(dat1$HC_Expenditure_Per_Capita, dat1$Cases_per_mill)
plot(dat1$Stay_Home_Req, dat1$Cases_per_mill)
plot(dat2$Life_Expectancy, dat2$Cases_per_mill)
plot(dat2$HC_Expenditure_Per_Capita, dat2$Cases_per_mill)
plot(dat2$Stay_Home_Req, dat2$Cases_per_mill)

par(mfrow=c(2,2))
plot(dat1$Gathering_Restrictions, dat1$Cases_per_mill)
plot(dat1$Internal_Movement_Restrict, dat1$Cases_per_mill)
plot(dat2$Gathering_Restrictions, dat2$Cases_per_mill)
plot(dat2$Internal_Movement_Restrict, dat2$Cases_per_mill)
plot(dat2$Tests_per_mill, dat2$Cases_per_mill)
```

# 3. Data Analysis
## 3.1. Attempt with a normal MLR
### 3.1.1. Cases_per_mill as dependent variable
```{r}
# fit everything into one big MLR
cases_mlr = lm(Cases_per_mill ~ Population + Population_Density + GDP_Per_Cap + Resp_Death_Rate + Median_Age + Contact_Tracing + INT_Travel_Controls + Physicians_Per_1000_ppl + Percent_Urbanization + Life_Expectancy + HC_Expenditure_Per_Capita + Stay_Home_Req + Gathering_Restrictions + Internal_Movement_Restrict + Tests_per_mill, data = dat2)
summary(cases_mlr)
par(mfrow=c(2,2))
plot(cases_mlr)
vif(cases_mlr)
```

### 3.1.2. Cases_per_mill as dependent variable, with life expectancy removed
```{r}
# Same as 3.1.1., but with life expectancy removed
cases_mlr = lm(Cases_per_mill ~ Population + Population_Density + GDP_Per_Cap + Resp_Death_Rate + Median_Age + Contact_Tracing + INT_Travel_Controls + Physicians_Per_1000_ppl + Percent_Urbanization + HC_Expenditure_Per_Capita + Stay_Home_Req + Gathering_Restrictions + Internal_Movement_Restrict + Tests_per_mill, data = dat2)
summary(cases_mlr)
par(mfrow=c(2,2))
plot(cases_mlr)
vif(cases_mlr)
```

### 3.1.3. Deaths_per_mill as dependent variable, with life expectancy removed
```{r}
# same as 3.1.2., but with deaths instead of cases as dependent variable
death_mlr = lm(Death_per_mill ~ Population + Population_Density + GDP_Per_Cap + Resp_Death_Rate + Median_Age + Contact_Tracing + INT_Travel_Controls + Physicians_Per_1000_ppl + Percent_Urbanization + HC_Expenditure_Per_Capita + Stay_Home_Req + Gathering_Restrictions + Internal_Movement_Restrict + Tests_per_mill, data = dat2)
summary(death_mlr)
par(mfrow=c(2,2))
plot(death_mlr)
vif(death_mlr)
```

## 3.2. Attempt with the GAM function
### 3.2.1. Cases_per_mill as dependent variable
```{r}
# seemingly linear variables had the 's' taken off.
# however, population was the only variable such that taking the 's' off resulted in weird behavior
# (such as adj R-squared decreasing significantly), so the 's' was left on for population
cases_GAM = gam(Cases_per_mill ~
                  s(Population) +
                  s(Population_Density) +
                  s(GDP_Per_Cap) +
                  Resp_Death_Rate +
                  s(Median_Age) +
                  Contact_Tracing +
                  s(INT_Travel_Controls) +
                  s(Physicians_Per_1000_ppl) +
                  s(Percent_Urbanization) +
                  s(HC_Expenditure_Per_Capita) +
                  Stay_Home_Req +
                  Gathering_Restrictions +
                  Internal_Movement_Restrict +
                  s(Tests_per_mill),
                data = dat2)
summary(cases_GAM)
par(mfrow=c(2,3))
plot(cases_GAM)
```


### 3.2.2. Cases_per_mill as dependent variable, two outliers removed
```{r}
# these are indices for India and Singapore
dat2_no_Outliers <- dat2[-c(39, 84),]
cases_GAM = gam(Cases_per_mill ~
                  s(Population) +
                  s(Population_Density) +
                  s(GDP_Per_Cap) +
                  Resp_Death_Rate +
                  s(Median_Age) +
                  Contact_Tracing +
                  s(INT_Travel_Controls) +
                  s(Physicians_Per_1000_ppl) +
                  s(Percent_Urbanization) +
                  s(HC_Expenditure_Per_Capita) +
                  Stay_Home_Req +
                  Gathering_Restrictions +
                  Internal_Movement_Restrict +
                  s(Tests_per_mill),
                data = dat2_no_Outliers)
summary(cases_GAM)
par(mfrow=c(2,3))
plot(cases_GAM)
```

### 3.2.3. Deaths_per_mill as dependent variable, two outliers removed
```{r}
# Same as 3.2.2., but with deaths instead of cases
death_GAM = gam(Death_per_mill ~
                  s(Population) +
                  s(Population_Density) +
                  s(GDP_Per_Cap) +
                  Resp_Death_Rate +
                  s(Median_Age) +
                  Contact_Tracing +
                  s(INT_Travel_Controls) +
                  s(Physicians_Per_1000_ppl) +
                  Percent_Urbanization +
                  s(HC_Expenditure_Per_Capita) +
                  s(Stay_Home_Req) +
                  Gathering_Restrictions +
                  s(Internal_Movement_Restrict) +
                  Tests_per_mill,
                data = dat2_no_Outliers)
summary(death_GAM)
par(mfrow=c(2,3))
plot(death_GAM)
```
